---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup-knitr, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# appm3310.final

This R package contains computation work for the APPM 3310 final project.

## Installation
To use the package, you first need to install it.

You can install the development version of appm3310.final like so:

```{r install-package, eval=FALSE}
library(remotes)
install_git("https://github.com/joshwlivingston/appm3310.final")
```
<br>

## Prepared data
After installation, you can load the cleaned data, which comes with the package.
```{r load-cleaned-data}
library(appm3310.final)
ncaam |> head()
```
<br>

## Data Prep
Alternatively, you can read the raw data and clean it using a provided helper function.

### Raw data 
The data was collected from 
[sports-reference.com](https://stathead.com/basketball/cbb/team-game-finder.cgi?request=1&comp_type=reg&game_status=1&order_by=date&match=team_game&year_max=2024&order_by_asc=1&timeframe=seasons&comp_id=NCAAM&year_min=2024). Currently, data for 
the 2023-24 NCAA men's basketball regular season has been collected. The data contains 
team-game level results for all NCAA men's basketball games. That is, there are two rows 
for each game in the data, one per team.You can call `read_ncaa_data(league = "NCAAM")` 
to read the raw data. 

If using `read_ncaa_data()`, you'll need to authenticate with your Google sheets account.
```{r read-data, eval=FALSE}
ncaam_raw <- read_ncaa_data(league = "NCAAM")
```

### Cleaned data
You will have to clean the data before using it. You can do so using `clean_ncaa_data()`.
```{r clean-data, eval=FALSE}
ncaam_cleaned <- ncaam_raw |> clean_ncaa_data()
```

## Prepare for computation
To prepare the data for ranking computation, you need to convert teh data to square 
matrix form. The functions in the `ranking_table_prep_*()` are the first step.

### Prepare ranking table
You can prepare the matrix for ranking using the direct win-loss method using 
`ranking_table_prep_direct_win_loss()`. This method assigns a 0 for a loss, 0.5 for a 
tie, and 1 for a win, then sums the results across games.
```{r prepare-data}
ncaam_prepared <- 
  ncaam |> 
  ranking_table_prep_direct_win_loss()
```
<br>

Looking at the teams that made it to the final four, you can validate regular season 
results. Use `view_subset()` to filter the matrix to specific teams.
```{r view-final-four}
library(dplyr)
final_four <- c("NC State", "Purdue", "Tennessee", "Duke")
ncaam_prepared |> view_subset(final_four)
```
<br>

### Convert table to matrix
Then, you'll have to convert the table to matrix form, which is done with the function 
`to_matrix()`.
```{r to-matrix}
ncaam_matrix <- 
  ncaam_prepared |> 
    to_matrix()

dim(ncaam_matrix)
```
<br>

### Matrix tests
`to_matrix()` calls two checks to ensure the data is ready for computation. These 
functions do not print anything; they simply error if needed to. These checks are 
performed automatically.
<br>

`check_is_square()` checks that the matrix has equal dimensions. 
```{r check-is-square}
ncaam_matrix |> check_is_square()
```
<br>

`check_nonnegative_matrix()` checks that all values in the matrix are >= 0
```{r check-nonnegative}
ncaam_matrix |> check_nonnegative_matrix()
```
<br>

## Computing eigenvalues
After preparing the matrix, you can compute eigenvalues. For the example above, we use 
the power method to approximate the eigenvalues. `eigenvalues_power_method()` defaults to 
1000 maximum iterations, and a convergence tolerance of 1e-10. These can be altered as 
arguments to the function. For more details, see `?eigenvalues_power_method`.
```{r compute-eigenvalues}
ncaam_eigenvalues <- 
  ncaam_matrix |> 
  eigenvalues_power_method()
```
<br>

`eigenvalues_power_method()` returns a list with two tibbles: the eigenvalues, and the 
results from the iteration.
```{r view-eigenvalues}
str(ncaam_eigenvalues)
```
<br>

To view the results from the iteration, you can call `plot_lambdas()`
```{r lambda-iteration-plot}
ncaam_eigenvalues |> plot_lambdas()
```
<br>

To compute the rankings, you can call `rankings_from_eigenvalues()`
```{r compute-rankings}
ncaam_eigenvalues |> 
  rankings_from_eigenvalues() |> 
  head(10)
```
<br><br>
